---
title: "Data Analysis of Movie Recommender Project"
date: "12/11/2023"
author: "Vivek Patel"
output:
  pdf_document: default
  html_notebook:
    theme: readable
    toc: yes
    toc_float: yes
  html_document:
    toc: yes
    df_print: paged
    keep_md: true
---


```{r, warning=FALSE}
# Load necessary libraries
library(dplyr)
library(ggplot2)
library(recommenderlab)
library(DT)
library(data.table)
library(reshape2)
```

## Download Data

```{r}
myurl = "https://raw.githubusercontent.com/Vivek2696/movie-recommender/main/data/"
```


Check the readme file to understand the format of the other three files. You can also download them from the original site: https://grouplens.org/datasets/movielens/.


## Read in data

### ratings data

```{r, warning=FALSE}
# use colClasses = 'NULL' to skip columns
ratings = read.csv(paste0(myurl, 'ratings.dat?raw=true'), 
                   sep = ':',
                   colClasses = c('integer', 'NULL'),
                   header = FALSE)
colnames(ratings) = c('UserID', 'MovieID', 'Rating', 'Timestamp')
head(ratings)
```

### movies data

```{r}
movies = readLines(paste0(myurl, 'movies.dat?raw=true'))
movies = strsplit(movies, split = "::", fixed = TRUE, useBytes = TRUE)
movies = matrix(unlist(movies), ncol = 3, byrow = TRUE)
movies = data.frame(movies, stringsAsFactors = FALSE)
colnames(movies) = c('MovieID', 'Title', 'Genres')
movies$MovieID = as.integer(movies$MovieID)

# convert accented characters
movies$Title[73]
movies$Title = iconv(movies$Title, "latin1", "UTF-8")
movies$Title[73]

# extract year
movies$Year = as.numeric(unlist(
  lapply(movies$Title, function(x) substr(x, nchar(x)-4, nchar(x)-1))))
```

```{r}
head(movies)
```


### user data

```{r}
users = read.csv(paste0(myurl, 'users.dat?raw=true'),
                 sep = ':', header = FALSE)
users = users[, -c(2,4,6,8)] # skip columns
colnames(users) = c('UserID', 'Gender', 'Age', 'Occupation', 'Zip-code')

head(users)
```

For users, Gender is denoted by “M” for male and “F” for female, Age is chosen from the following ranges:
```
*  1:  "Under 18"
* 18:  "18-24"
* 25:  "25-34"
* 35:  "35-44"
* 45:  "45-49"
* 50:  "50-55"
* 56:  "56+"
```
and Occupation is chosen from the following choices:
```
*  0:  "other" or not specified
*  1:  "academic/educator"
*  2:  "artist"
*  3:  "clerical/admin"
*  4:  "college/grad student"
*  5:  "customer service"
*  6:  "doctor/health care"
*  7:  "executive/managerial"
*  8:  "farmer"
*  9:  "homemaker"
* 10:  "K-12 student"
* 11:  "lawyer"
* 12:  "programmer"
* 13:  "retired"
* 14:  "sales/marketing"
* 15:  "scientist"
* 16:  "self-employed"
* 17:  "technician/engineer"
* 18:  "tradesman/craftsman"
* 19:  "unemployed"
* 20:  "writer"
```

## Exploration

### Check for unique users

```{r}
dim(users)
length(unique(ratings$UserID))
```


### Check for unique movies

```{r}
dim(movies)
length(unique(ratings$MovieID))
```

```{r}
movies_not_rated = movies %>% 
  filter(!(MovieID %in% ratings$MovieID))
dim(movies_not_rated)
```

### Dist of ratings

Most of the ratings are in the 3-5 range.

```{r}
tmp = data.frame(Rating = 1:5, 
                 freq = as.vector(table(ratings$Rating)/nrow(ratings)))
ggplot(data = tmp, aes(x = Rating, y = freq)) +
  geom_bar(stat="identity", fill = 'steelblue', width = 0.6) + 
  geom_text(aes(label=round(freq, dig=2)), 
                vjust=1.6, color="white", size=3.5) +
  theme_minimal()
```

### Ratings per users

```{r}
tmp = ratings %>% 
  group_by(UserID) %>% 
  summarize(ratings_per_user = n()) 
summary(tmp$ratings_per_user)
stem(tmp$ratings_per_user)
sum(tmp$ratings_per_user > 500)
sort(tmp$ratings_per_user[tmp$ratings_per_user>1300])
```

```{r}
tmp %>%
  ggplot(aes(ratings_per_user)) +
  geom_bar(fill = "steelblue") + coord_cartesian(c(20, 500))
```

Combining users and tmp, you could further explore how ratings_per_user depends on Gender, Age, and Occupation of users.

```{r}
tmp = tmp %>% full_join(users, by = 'UserID')
```

### Ratings per movie

There are 31 movies that have received more than 2000 ratings. The most popular movie is “American Beauty (1999)”, followed by the “Star Wars” series. Throughout, popular means receiving many ratings; a popular movie may not be a highly-rated movie.

```{r}
tmp = ratings %>% 
  group_by(MovieID) %>% 
  summarize(ratings_per_movie = n(), ave_ratings = mean(Rating)) %>%
  inner_join(movies, by = 'MovieID')
summary(tmp$ratings_per_movie)
```

```{r}
tmp %>% 
  filter(ratings_per_movie > 2000) %>%
  arrange(desc = ratings_per_movie) %>%
  select(c("Title", "ratings_per_movie")) %>%
  print(n = 31)
```

```{r}
tmp %>% ggplot(aes(ratings_per_movie)) + 
  geom_bar(fill = "steelblue", width = 1) + coord_cartesian(c(1,1500))
```

The top ten highly-rated (based on their average ratings) among all movies that have received at least 1000 ratings.

```{r}
small_image_url = "https://github.com/Vivek2696/movie-recommender/blob/main/data/MovieImages/"
ratings %>% 
  group_by(MovieID) %>% 
  summarize(ratings_per_movie = n(), 
            ave_ratings = round(mean(Rating), dig=3)) %>%
  inner_join(movies, by = 'MovieID') %>%
  filter(ratings_per_movie > 1000) %>%
  top_n(10, ave_ratings) %>%
  mutate(Image = paste0('<img src="', 
                        small_image_url, 
                        MovieID, 
                        '.jpg?raw=true"></img>')) %>%
  select('Image', 'Title', 'ave_ratings') %>%
  arrange(desc(-ave_ratings)) %>%
  datatable(class = "nowrap hover row-border", 
            escape = FALSE, 
            options = list(dom = 't',
                          scrollX = TRUE, autoWidth = TRUE))
```


### Dist of Genres

First cretae a bibary indicator for the 18 genres for each movie.

```{r}
genres = as.data.frame(movies$Genres, stringsAsFactors=FALSE)
tmp = as.data.frame(tstrsplit(genres[,1], '[|]',
                              type.convert=TRUE),
                    stringsAsFactors=FALSE)
genre_list = c("Action", "Adventure", "Animation", 
               "Children's", "Comedy", "Crime",
               "Documentary", "Drama", "Fantasy",
               "Film-Noir", "Horror", "Musical", 
               "Mystery", "Romance", "Sci-Fi", 
               "Thriller", "War", "Western")
m = length(genre_list)
genre_matrix = matrix(0, nrow(movies), length(genre_list))
for(i in 1:nrow(tmp)){
  genre_matrix[i,genre_list %in% tmp[i,]]=1
}
colnames(genre_matrix) = genre_list
remove("tmp", "genres")
```

Then we can output historograms of the 18 genres baesd on movies or based on ratings.

```{r}
data.frame(Genres = genre_list, 
                 Freq = as.vector(colMeans(genre_matrix))) %>% 
  ggplot(aes(reorder(Genres, Freq), Freq, fill = Freq)) + 
  geom_bar(stat = "identity") + 
    geom_text(aes(label = round(Freq, dig=2)), 
            position = position_stack(vjust = 0.5), 
            color="white", size=3) + 
  coord_flip() + 
  scale_colour_brewer(palette="Set1") + 
  labs(y = 'Frequency', x = 'Genre')
```

```{r}
tmp = ratings %>% 
  left_join(data.frame(MovieID = movies$MovieID, genre_matrix), 
            by = "MovieID") %>%
  select(-c("UserID", "MovieID", "Rating", "Timestamp"))
data.frame(Genres = genre_list, 
                 Popularity = as.vector(colMeans(tmp))) %>% 
  ggplot(aes(reorder(Genres, Popularity), Popularity, fill = Popularity)) + 
  geom_bar(stat = "identity") + 
  geom_text(aes(label = round(Popularity, dig=3)), 
            position = position_stack(vjust = 0.5), 
            color="white", size=3) + 
  coord_flip() + 
  labs(y = 'Popularity', x = 'Genre')
```

About half of the movies belong to only one genre; about one-third are categorized into two genres; a few are categorized into more than four genres. The movie “The Transformers” (1986) are categorized into six genres: Action, Animation, Children’s, Sci-Fi, Thriller, and War.

```{r}
tmp = rowSums(genre_matrix)
summary(tmp)

movies[which(tmp==6), ]
movies[which(tmp==5), ]
```


For illustration purpose only, let’s assume movies contains all the movies available to users from 1919 to 2000. Then we can compute the cumulative percentages of the 18 genres over year from 1919 to 2000 and store them in the 81-by-19 matrix tmp. For example, till 2000, users can access about 7.8% Action, 4.4% Adventure, 25% Drama, etc. A graphical display of such CDF over 10 generes are displayed below.

```{r}
# range(movies$Year) % 1919 to 2000
tmp = data.frame(Year = movies$Year, genre_matrix) %>%
  group_by(Year) %>%
  summarise_all(sum)
tmp[,-1] = apply(tmp[, -1], 2, cumsum)
tmp[,-1] = tmp[,-1]/sum(tmp[nrow(tmp), -1])
print(round(tmp[nrow(tmp),-1], dig=3))
```

```{r}
tmp = reshape2::melt(tmp, id.vars="Year") 
tmp %>%
  ggplot(aes(Year, value, group = variable)) +
  geom_area(aes(fill = variable)) + 
  geom_line(aes(group = variable), position = "stack")
```











